---
title: "Reproducible Research Course"
author: "Remzi Kurshumliu"
date: "21 August 2016"
output: pdf_document
---
 

## Entry

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. 


```{r,echo = TRUE,include=FALSE}  

## ----loaddata------------------------------------------------------------
require(lattice) 
library(data.table)
library(ggplot2)

#First I want to make sure that default directory is set right before I start with the work
setwd("C:\\RSWeek2")
# we define name of the file in unzipped and zipped format and check if unzipped format exists. If not, we unzip it'
file = "activity.csv"
zipfilename = "repdata_data_activity.zip"
if (!file.exists(file)) {
  unzip(zipfilename)
}

dataset <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))

dataset$date <- as.Date(dataset$date, format = "%Y-%m-%d")
dataset$interval <- as.factor(dataset$interval)

str(dataset)
```

# What is mean total number of steps taken per day?
```{r,echo = TRUE}
steps_per_day <- aggregate(steps ~ date, dataset, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)

ggplot(steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "orange", binwidth = 1000) + 
  labs(title="Histogram of Steps Taken per Day", 
       x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 

steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)

```

#What is the average daily activity pattern?

```{r,echo = TRUE}
steps_per_interval <- aggregate(dataset$steps, 
                                by = list(interval = dataset$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers
##this helps in plotting
steps_per_interval$interval <- 
  as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")

ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
  geom_line(color="orange", size=1) +  
  labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
  theme_bw()

max_interval <- steps_per_interval[which.max(  
  steps_per_interval$steps),]
#Imputing missing values
```
# Number of missing values
```{r,echo = TRUE}
missing_vals <- sum(is.na(dataset$steps))

```
# Strategy for dealimng with the missing values
```{r,echo = TRUE}

na_fill <- function(data, pervalue) {
  na_index <- which(is.na(data$steps))
  na_replace <- unlist(lapply(na_index, FUN=function(idx){
    interval = data[idx,]$interval
    pervalue[pervalue$interval == interval,]$steps
  }))
  fill_steps <- data$steps
  fill_steps[na_index] <- na_replace
  fill_steps
}

dataset_fill <- data.frame(  
  steps = na_fill(dataset, steps_per_interval),  
  date = dataset$date,  
  interval = dataset$interval)
str(dataset_fill)

sum(is.na(dataset_fill$steps))
```

# A histogram of the total number of steps taken each day
```{r,echo = TRUE}
fill_steps_per_day <- aggregate(steps ~ date, dataset_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##Code to plot the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "blue", binwidth = 1000) + 
  labs(title="Histogram of Steps Taken per Day", 
       x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
#Calculate and report the mean and median total number of steps taken per day.

```
# Calculate as well as report  mean and median total number of steps taken per day.
``` {r,echo = TRUE}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```

# What is the impact of imputing missing data on the estimates of the total daily number of steps?
It can be concluded that  that while the mean value remains unmodified, the median value did shift and virtual matches to the mean.
 
# Are there differences in activity patterns between weekdays and weekends?
```{r,echo = TRUE}
weekdays_steps <- function(data) {
  weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                              FUN=mean, na.rm=T)
  # convert to integers for plotting
  weekdays_steps$interval <- 
    as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
  colnames(weekdays_steps) <- c("interval", "steps")
  weekdays_steps
}

data_by_weekdays <- function(data) {
  data$weekday <- 
    as.factor(weekdays(data$date)) # weekdays
  weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
  weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))
  
  weekend_steps <- weekdays_steps(weekend_data)
  weekday_steps <- weekdays_steps(weekday_data)
  
  weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
  weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
  
  data_by_weekdays <- rbind(weekend_steps, weekday_steps)
  data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
  data_by_weekdays
}

data_weekdays <- data_by_weekdays(dataset_fill)

ggplot(data_weekdays, aes(x=interval, y=steps)) + 
  geom_line(color="violet") + 
  facet_wrap(~ dayofweek, nrow=2, ncol=1) +
  labs(x="Interval", y="Number of steps") +
  theme_bw()

```
It can be seen from the graph presented abve  above that activity on the weekday has the greatest peak from all steps intervals. But, we can see too that weekends activities has more peaks over a hundred than weekdays do. 